On the Social Compatibility of Genome Analysis

### 2022-06-16 draft.tex
formatting and final changes

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{chronosys}
\usepackage{textcomp}
\usepackage{xcolor}

@@

@@

To recapitulate, this approximate timeline in Figure 1 should give an overview
\vspace{0.5cm}
	\startchronology[startyear=1920, stopdate=2022,arrowheight=0.2cm,arrowwidth=0.2cm,width=8cm,height=0.05cm]
	\chronoevent{2022}{This Paper}
	\chronoevent[markdepth=120pt]{2011}{Successful CRISPR Application}
	\chronoevent[markdepth=80pt]{2003}{HGP Completion}
	\chronoevent[markdepth=40pt]{1996}{Cloning of Dolly}
	\chronoevent{1987}{CRISPR Discovery}
	\chronoevent[markdepth=40pt]{1944}{A-Ml-Mc Experiment}
	\chronoevent{1928}{Griffiths Experiment}
	\stopchronology
It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non-modified cells\cite{crispr}.
This can be worked around in three ways.
\begin{itemize}
\item Modify a sufficient number of cells (most likely infeasible for broad changes)
\item Modify in an early stage of development, where only a number of cells exist
\item Germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
\end{itemize}
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

Advances here have sprouted the field of pharmagenomics, which utilizes genetic information to practice precision medicine.
Just like ethnicity, genetic profile has been shown to affect drug response, thus using this knowledge to personalize prescriptions can increase effectiveness and lower cost and side effects \cite{costsaving}.
Predictive genetic testing does not come without concerns from both experts and the public, the most prominent ones will be discussed in the following:

\subsubsection{Patient Autonomy}
The primary role of any new medical instrument should be to serve, therefore patient autonomy and informed consent are non-negotiable, which necessitates public education.
Broad, as well as targeted at populations most susceptible to specific genetic conditions.
Equally as important is access to testing facilities themselves as well as short wait times for both test and treatment\cite{publicopinion}.

@@

Overall the medical sectors holds much potential for the use of gene testing and therapy, with cheaper and more effective treatment, more reliable diagnoses and hope of a cure for previously terminal disease.

\section{Consequences for Society: Social Compatibility}
Genetic testing has already established itself in the developed world. This chapter deals with the expected consequences of widespread application of analysis and one day manipulation tools, as well as the commercial exploitation of the underlying technology.
\subsection{Analysis}
Analogous to any piece of sensitive and non-public information, knowledge derived from genetic analysis is prone to being exploited as soon as conflicting interest arises.
Companies refusing to employ people who lost part of the genetic lottery would undoubtedly be unfair.
Even more so with technology being fallible. Why should someone whose condition is yet to be diagnosed be more fit for a job than someone with a genetic condition that can already be tested for?

@@

Despite considerable efforts over the past several decades and major progress toward unraveling the genetic etiology of personality traits, the genetic variants that influence personality are only beginning to be identified\cite{geneticspersonality}. 
The genetic testing provider 23andMe divides their possible personality traits into three report groups. 
\begin{itemize}
\item The first report is about "Physical Features" which includes information about Hair Photobleaching, Freckles, Skin Pigmentation, and more. 
\item The secound report focuses on "Taste and Smell" and reveals more about the customers Cilantro Taste Aversion, Sweet vs. Salty, Asparagus Odor Detection, and. 
\item The third report is called "Weird and wonderful" and is about Photic Sneeze Reflex, Ability to Match Musical Pitch, Misophonia (hatred of the sound of chewing), and more\cite{23personality}.
\end{itemize}
Although a large number of personality traits are listed and research into genetic personality research is still in its infancy, there is no reference to the accuracy of the test results on the provider's website. Unfortunately, this is not an isolated case of lack of transparency by a DNA testing provider.

@@

Laestadius, Rich \& Auer investigated whether 30 different direct-to-consumer genetic testing websites followed the guidelines created by professional and government bodies to promote transparency in two aspects.
\begin{itemize}
\item Transparency regarding confidentiality and privacy.
\item Use of data
\end{itemize}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}

\usepackage{enumitem}

@@

\begin{figure}[h!]
	\includegraphics[width=\textwidth/2]{timeline.png}
	\caption{A rough timeline since Griffiths Experiment}
	\label{timeline}
\end{figure}

@@

To recapitulate, figure \ref{timeline} should give an overview.

It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non-modified cells\cite{crispr}.
This can be worked around in three ways.
\begin{itemize}[leftmargin=*]
\item Modify a sufficient number of cells (most likely infeasible for broad changes)
\item Modify in an early stage of development, where only a number of cells exist
\item Germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
\end{itemize}
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

Advances here have sprouted the field of pharmagenomics, which utilizes genetic information to practice precision medicine.
Just like ethnicity, genetic profile has been shown to affect drug response, thus using this knowledge to personalize prescriptions can increase effectiveness and lower cost and side effects \cite{costsaving}.
Predictive genetic testing does not come without concerns from both experts and the public, the most prominent ones will be discussed in the following:
{\parindent0pt
\subsubsection{Patient Autonomy}
The primary role of any new medical instrument should be to serve, therefore patient autonomy and informed consent are non-negotiable, which necessitates public education.
Broad, as well as targeted at populations most susceptible to specific genetic conditions.
Equally as important is access to testing facilities themselves as well as short wait times for both test and treatment\cite{publicopinion}.

@@

Overall the medical sectors holds much potential for the use of gene testing and therapy, with cheaper and more effective treatment, more reliable diagnoses and hope of a cure for previously terminal disease.
}
\section{Consequences for Society: Social Compatibility}
Genetic testing has already established itself in the developed world. This chapter deals with the expected consequences of widespread application of analysis and one day manipulation tools, as well as the commercial exploitation of the underlying technology.
\subsection{Analysis}
Analogous to any piece of sensitive and non-public information, knowledge derived from genetic analysis is prone to being exploited as soon as conflicting interest arises.
Companies refusing to employ people who lost part of the genetic lottery would undoubtedly be unfair.
Even more so with technology being fallible. Why should someone whose condition is yet to be diagnosed be more fit for a job than someone with a genetic condition that can already be tested for?

@@

Despite considerable efforts over the past several decades and major progress toward unraveling the genetic etiology of personality traits, the genetic variants that influence personality are only beginning to be identified\cite{geneticspersonality}. 
The genetic testing provider 23andMe divides their possible personality traits into three report groups. 
\begin{itemize}[leftmargin=*]
\item The first report is about "Physical Features" which includes information about Hair Photobleaching, Freckles, Skin Pigmentation, and more. 
\item The secound report focuses on "Taste and Smell" and reveals more about the customers Cilantro Taste Aversion, Sweet vs. Salty, Asparagus Odor Detection, and. 
\item The third report is called "Weird and wonderful" and is about Photic Sneeze Reflex, Ability to Match Musical Pitch, Misophonia (hatred of the sound of chewing), and more\cite{23personality}.
\end{itemize}
Although a large number of personality traits are listed and research into genetic personality research is still in its infancy, there is no reference to the accuracy of the test results on the provider's website. Unfortunately, this is not an isolated case of lack of transparency by a DNA testing provider.

@@

Laestadius, Rich \& Auer investigated whether 30 different direct-to-consumer genetic testing websites followed the guidelines created by professional and government bodies to promote transparency in two aspects.
\begin{itemize}[leftmargin=*]
\item Transparency regarding confidentiality and privacy.
\item Use of data
\end{itemize}
